Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm':' A Case Study

### 2022-04-15 Paper_Draft2/Draft2.Rmd
Tense edits

The data we are using for this project is the COMPAS General Recidivism Risk Scores data set from the AI Fairness 360 (AIF360) toolkit [@aif360-oct-2018], which does the same initial pre-processing as ProPublica. The raw data has 6,167 rows and each row represents an arrest charge for a defendant. AIF360’s COMPAS data includes the defendant's age, race, sex, what they were charged with, and whether or not the defendant ultimately recidivated within a two-year period after their arrest. For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants versus Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American. Our data therefore has 5,723 rows (Figure \ref{fig:table snip}), with the below distributions of race (Figure \ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot} and Table \ref{tab:recid table}).

```{r table snip, echo = FALSE, fig.align='center', fig.cap= 'This table is a snippet of the data set we will be using.', out.width= "100%"}
knitr::include_graphics("../images/table_snippet.png")
```

```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant races.', out.width= "100%"}
knitr::include_graphics("../images/race_bar_plot_new.png")
```

```{r age plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of defendant ages by race.', out.width= "100%"}
knitr::include_graphics("../images/age_race_plot_new.png")
```

```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes.', out.width= "100%"}
knitr::include_graphics("../images/recid_bar_plot_new.png")
```

```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes by race.', out.width= "100%"}
knitr::include_graphics("../images/race_recid_bar_plot_new.png")
```

```{r, include = FALSE}
library(tidyverse)
```

```{r, include = FALSE}
x <- tribble(~`Two Year Recidivism by Race`, ~`Recidivated`, ~`Survived`,
             "African American", "1661", "1512",
             "Caucasian", "822", "1278")
```

```{r, echo = FALSE}
knitr::kable(x, caption = "This table shows the distribution of recidivism by race \\label{tab:recid table}")
```

@@

For the next step of our experiment, we needed to determine where in the data science pipeline we can mitigate the most bias, using pre-processing, in-processing, and post-processing de-biasing algorithms. These are all based on using predictive models to figure out how we can “fix” the bias that is present. 
Pre-processing refers to assessing the training data, and it is the most flexible method because it has not yet trained a model that may carry assumptions about the data. It is important to keep in mind that pre-processing prevents assumptions in the modeling, but does not account for the bias in data collection. With pre-processing, we use the re-weighing algorithm from AIF360 which weights the examples in each (group, label) combination differently to ensure fairness before classification [@aif360-oct-2018]. We used a logistic regression model for this algorithm, as it is the easiest to interpret in the given context. After running the fairness metrics using the pre-processing algorithm, we were able to compare our results to the baseline metrics from the previous section.

Another approach, in-processing, explores how bias can be targeted while building a model. It assesses how we build learning algorithms, which often prioritize accuracy over fairness. The technique we use is the Meta Classification algorithm, which accounts for the fairness metric as part of the input and returns a classifier optimized by that particular metric. In order to do this, we first needed to convert our data frame into a data type called a . Similarly to pre-processing, we compared the results of our in-processing methods with both the baseline and the pre-processing to gauge which method so far has better individual or group fairness.

Our last approach, post-processing, makes transformations on model outputs. Similarly to the pre-processing, it is flexible because it does not work directly with the algorithm, but instead has direct access to all the predicted classification values. Because the algorithms are often so complex and difficult to interpret, post-processors are trained in a black box. To get around this, we use the PostProcessingMeta class from AIF360 which combines the training and prediction for a random estimator and the post-processor while simultaneously splitting the data set. Based on these results, we looked comparatively at all the approaches to determine which we assessed to be the most fair.

>>>>

The data we are using for this project is the COMPAS General Recidivism Risk Scores data set from the AI Fairness 360 (AIF360) toolkit [@aif360-oct-2018], which does the same initial pre-processing as ProPublica. The raw data has 6,167 rows and each row represents an arrest charge for a defendant. AIF360’s COMPAS data includes the defendant's age, race, sex, what they were charged with, and whether or not the defendant ultimately recidivated within a two-year period after their arrest. For the purposes of our project, which endeavors to evaluate the differing effects of the COMPAS algorithm on white defendants versus Black defendants, we have filtered the data to only include individuals whose race is listed as Caucasian or African-American. Our data therefore has 5,723 rows (Figure \ref{fig:table snip}), with the below distributions of race (Figure \ref{fig:race plot}), age (Figure \ref{fig:age plot}), and two year recidivism rate (Figures \ref{fig:recid plot} & \ref{fig:recid race plot}).

```{r table snip, echo = FALSE, fig.align='center', fig.cap= 'This table is a snippet of the data set we will be using.', out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/table_snippet_new.png")
```

```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'The race distribution for black people is 3000 and the distribution for white people is 2100. Therefore, there are more black people in the dataset than white people.', out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/race_bar_plot.png")
```

```{r age plot, echo=FALSE, fig.align='center', fig.cap= "This plot is a Kernel Density Estimate (KDE) plot. The y-axis represents the KDE of the Probability Distribution Function of the age variable in our dataset. The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points. The purple dotted line shows the average age of Black defendants and the green dotted line shows the average age of white defendants. For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.", out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/race_age_plot.png")
```

```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'Of both black and white people, there is a higher percentage of people who did not recidivate than who did. 2500 people recidivated, whereas approximately 2600 did not.', out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/recid_bar_plot.png")
```

```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants. Black defendants recidivate are more likely than not to recidivate. However, we can also see that there are more Black defendants in the dataset in general.', out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/race_recid_plot.png")
```

@@

For the next step of our experiment, we need to determine where in the data science pipeline we can mitigate the most bias, using pre-processing, in-processing, and post-processing de-biasing algorithms. These are all based on using predictive models to figure out how we can “fix” the bias that is present. 
Pre-processing refers to assessing the training data, and it is the most flexible method because it has not yet trained a model that may carry assumptions about the data. It is important to keep in mind that pre-processing prevents assumptions in the modeling, but does not account for the bias in data collection. With pre-processing, we use the re-weighing algorithm from AIF360 which weights the examples in each (group, label) combination differently to ensure fairness before classification [@aif360-oct-2018]. We use a logistic regression model for this algorithm, as it is the easiest to interpret in the given context. After running the fairness metrics using the pre-processing algorithm, we are able to compare our results to the baseline metrics from the previous section.

Another approach, in-processing, explores how bias can be targeted while building a model. It assesses how we build learning algorithms, which often prioritize accuracy over fairness. The technique we use is the Meta Classification algorithm, which accounts for the fairness metric as part of the input and returns a classifier optimized by that particular metric. Similarly to pre-processing, we compare the results of our in-processing methods with both the baseline and the pre-processing to gauge which method so far has better individual or group fairness.

Our last approach, post-processing, makes transformations on model outputs. Similarly to the pre-processing, it is flexible because it does not work directly with the algorithm, but instead has direct access to all the predicted classification values. Because the algorithms are often so complex and difficult to interpret, post-processors are trained in a black box. To get around this, we use the PostProcessingMeta class from AIF360 which combines the training and prediction for a random estimator and the post-processor while simultaneously splitting the data set. Based on these results, we look comparatively at all the approaches to determine which we assess to be the most fair.
